---
title: "Making Graphs"
output: html_document
vignette: >
  %\VignetteIndexEntry{Making graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)
```

```{r startupcode}

library(lehmansociology)
education_and_poverty<-create_educ_poverty_data()
library(ggplot2)

```
In SOC 345 we are using the graphing system known as ggplot2.

Ggplot is a system in which you start a plot and then layer additional features by adding 
commands using a +.

Here are some examples.  Note that right now they will not run because the data set is not created.


##Dotplot
Creating a plot always starts with a  ggplot()  command.
A dotplot has only one variable, the x. 
We are using the education_and_poverty dataset.
The scale_y_continuous helps to make the dotplot look better.

Every ggplot has a geom_  that says what kind of graph we are making.  In this case it is a dot plot.

````{r dotplot}
# Dot plot
# This makes the basic plot
ggplot(education_and_poverty,
       aes(x=PCTPOVALL_2013),
       scale_y_continuous(breaks=NULL)) +
  geom_dotplot(binwidth=1) 

````

But then we can add more features like labels and a title using the +

````{r dotplot2}
# Dot plot
ggplot(education_and_poverty,
       aes(x=PCTPOVALL_2013),
       scale_y_continuous(breaks=NULL)) +
  geom_dotplot(binwidth=1) +
  geom_rug() +
  labs( x="Percent in Poverty") +
  scale_y_continuous(breaks=NULL) +
  ggtitle("Fig # :")

````

## Histogram

Another basic plot type is the histogram. 

````{r histogram}
# Histogram

ggplot(education_and_poverty, aes(x=PCTPOVALL_2013)) +
  geom_histogram(binwidth=1) +
  labs(y = "Count",     x="Percent in Poverty") +
  ggtitle("Fig # :")

````

## Bar plots
Bar plots have two variables, the  variable defines the bars, and the y variable determines how tall the bars are.
In bar charts the x variable is discrete, so notice that there is a space between the bars.
```{r label="barplot"}

ggplot(education_and_poverty, 
       aes(x= reorder(Area_Name, PCTPOVALL_2013), y = PCTPOVALL_2013)) +
  geom_bar() +
    #Extras
  labs(y = "Percent in Poverty",     x="State") +
     # This line is going to rotate the state names by 90 degrees. You can try other values.
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Fig # :")

```


## Additions

ggplot will let us add lots of features. 
For example you can add vertical and horizontal lines to your plots to help make the more understandable.
Adding the lines below will add a vertical line (vline) at the mean and median.

   geom_vline(aes ( xintercept  = median(education_and_poverty$PCTPOVALL_2013)))
 
   geom_vline(aes ( xintercept  = mean(education_and_poverty$PCTPOVALL_2013)))  


## Maps

Here is the basic ggplot code for making a map.

This is a little more complex because you need to set up the data first.
```{r setupmaps}
# First we need to load an extra library
library(maps)
states_map <- map_data("state")

# We'll use this library if we want to play with colors in the maps
library("RColorBrewer")

```
If you look at the states_map data you will see that the state names are in lower case.
Unfortunately our Area_Name variable is in upper case so ... it won't match.  So first we have to deal with that and then we can merge the map data and our education_and_poverty data together.

```{r mapdata}

education_and_poverty$state <- tolower(education_and_poverty$Area_Name)

education_and_poverty_map <- merge(states_map, education_and_poverty, 
                                 by.x = "region", by.y="state")

```


This actually makes the map, and you an see that this part is very similar to every other ggplot.

```{r createemap}
# If you want to try colors uncomment the next line and change to the color you want.
# Also uncomment the scale_fill_gradientn line.
# You can change the color and the number of colors.

#colors<-brewer.pal(7,"Greens")
ggplot(education_and_poverty_map, aes(x=long, y = lat, group = group, 
                                      map_id = region, fill = PCTPOVALL_2013)) +
  geom_map(map = education_and_poverty_map, color = "black" ) +
  #scale_fill_gradientn(colors=colors) +
  coord_map("polyconic") +
  ggtitle("Fig #: ")

```


