---
title: "Making Tables"
author: "Elin Waring"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{UTF-8}
---

```{r, include = FALSE, message=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

library('lehmansociology')
library(dplyr) 
library(xtable)
education_and_poverty<-create_educ_poverty_data()
```

There are many kinds of "tables" that we make in presenting statistical data and doing statistical analysis.  A table can be anything that is organized into columns and rows.

For example you could consider presenting the summary statistics for a variable to be a table.
```{r, echo=FALSE, message=FALSE}

summary(poverty.states$PCTPOV05_2013)

```

However, usually we would make a table with 


```{r regions_table}

table(education_and_poverty$region)
# You can display this as proportions. Notice how the function builds from the inside out.
prop.table(table(education_and_poverty$region))

# Alternatively you could do this
mytable <- table(education_and_poverty$region)
prop.table(mytable)


```

Now we will do the same comparison (of means and medians) for different groups with a table of values.

There are many ways to do this, but this way uses some functions from the dplyr package and the magritte operator (%>%).

```{r comparison_table}
#start with your dataset 
education_and_poverty %>%
    #group it according the the variable you want
    group_by(region) %>% 
    # Use summarize to define the table columns
    # You can add any othe statistics you want, such as max() or min()
    # Add the new statistics to the list inside the parentheses.
    summarize(
            'Mean poverty' = mean(PCTPOVALL_2013), 
            'Median poverty' = median(PCTPOVALL_2013), 
            'IQR poverty' = IQR(PCTPOVALL_2013)
            )

```

