---
title: "Making Graphs"
output: html_document
vignette: >
  %\VignetteIndexEntry{Making graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ggplot2)
```

```{r startupcode, message=FALSE, warning=FALSE}

library(lehmansociology)
education_and_poverty<-create_educ_poverty_data()
library(ggplot2)

```
In SOC 345 we are using the graphing system known as ggplot2.

Ggplot is a system in which you start a plot and then layer additional features by adding 
commands using a +.

Here are some examples.  Note that right now they will not run because the data set is not created.


##Dotplot
Creating a plot always starts with a  ggplot()  command.
A dotplot has only one variable, the x. 
We are using the education_and_poverty dataset.
The scale_y_continuous helps to make the dotplot look better.

Every ggplot has a geom_  that says what kind of graph we are making.  In this case it is a dot plot.

````{r dotplot}
# Dot plot
# This makes the basic plot
ggplot(education_and_poverty,
       aes(x=PCTPOVALL_2013),
       scale_y_continuous(breaks=NULL)) +
  geom_dotplot(binwidth=1) 

````

But then we can add more features like labels and a title using the +

````{r dotplot2}
# Dot plot
ggplot(education_and_poverty,
       aes(x=PCTPOVALL_2013),
       scale_y_continuous(breaks=NULL)) +
  geom_dotplot(binwidth=1) +
  geom_rug() +
  labs( x="Percent in Poverty") +
  scale_y_continuous(breaks=NULL) +
  ggtitle("Fig # :")

````

## Histogram

Another basic plot type is the histogram. 
This also only has one variable and uses geom_histogram().

````{r histogram}
# Histogram

ggplot(education_and_poverty, aes(x=PCTPOVALL_2013)) +
  geom_histogram(binwidth=1) +
  labs(y = "Count",     x="Percent in Poverty") +
  ggtitle("Fig # :")

````

## Bar plots
Bar plots have two variables, the  variable defines the bars, and the y variable determines how tall the bars are.
In bar charts the x variable is discrete, so notice that there is a space between the bars.
Note we use geom_bar().

```{r barplot}

ggplot(education_and_poverty, 
       aes(x= Area_Name, PCTPOVALL_2013, y = PCTPOVALL_2013)) +
  geom_bar(stat="identity") +
    #Extras
  labs(y = "Percent in Poverty",     x="State") +
     # This line is going to rotate the state names by 90 degrees. You can try other values.
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Fig # :")

```

Here's a slightly different way that illustrates using advanced features, in this case sorting the values from low to high.
```{r barplot2 }

ggplot(education_and_poverty, 
       aes(x= reorder(Area_Name, PCTPOVALL_2013), y = PCTPOVALL_2013)) +
  geom_bar(stat="identity") +
  labs(y = "Percent in Poverty",     x="State") +
     # This line is going to rotate the state names by 90 degrees. You can try other values.
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Fig # :")

```


## Scatterplots


Scatter plots are used when you have two interval variables.
Notice that here we use geom_point() to indicate that we want a scatterplot, which is made up of points. 

```{r scatterplot}
ggplot(education_and_poverty,
       aes(x=Percent.of.adults.with.less.than.a.high.school.diploma..2009.2013,
           y=PCTPOVALL_2013)) +
  geom_point() +
  ggtitle("Fig # : Poverty Rate and High School Completion (for States)") +
  labs(x = "Percent of adults with less than highschool diploma",  
       y="Percent of population in poverty") 
```

## Box Plots
A box plot shows: 

 * The median (50th percentile, second quartile) (dark line in the middle)
 * The first quartile (25th percentile) (low value of box)
 * The third quartile (75th percentile) (high value of box of box)
 * The interquartile range is the distance from the high edge to the low edge
 * Lines showing the location of the farthest point that is still within  1.5 * the Interquartile above or below the boxes.
 * Points for individual outliers that are beyond the lines. These are usually considered outliers.
 
The x variable represents the groups and the y variable is the one we are comparing across groups.
 
```{r boxplots}

# In R, the term factor is used for nominal variables.
ggplot(education_and_poverty, 
       aes(x=factor(region), y=PCTPOVALL_2013)) +
 geom_boxplot()

```

## Additions

ggplot will let us add lots of features. 
For example you can add vertical and horizontal lines to your plots to help make the more understandable.
Adding the lines below will add a vertical line (vline) at the mean and median.

   geom_vline(aes ( xintercept  = median(education_and_poverty$PCTPOVALL_2013)))
 
   geom_vline(aes ( xintercept  = mean(education_and_poverty$PCTPOVALL_2013)))  


## Maps

Here is the basic ggplot code for making a map.

This is a little more complex because you need to set up the data first.
```{r setupmaps}
# First we need to load an extra library
library(maps)
states_map <- map_data("state")

# We'll use this library if we want to play with colors in the maps
library("RColorBrewer")

```
If you look at the states_map data you will see that the state names are in lower case.
Unfortunately our Area_Name variable is in upper case so ... it won't match.  So first we have to deal with that and then we can merge the map data and our education_and_poverty data together.

```{r mapdata}

education_and_poverty$state <- tolower(education_and_poverty$Area_Name)

education_and_poverty_map <- merge(states_map, education_and_poverty, 
                                 by.x = "region", by.y="state")

```


This actually makes the map, and you an see that this part is very similar to every other ggplot.

```{r createemap}
# If you want to try colors uncomment the next line and change to the color you want.
# Also uncomment the scale_fill_gradientn line.
# You can change the color and the number of colors.

#colors<-brewer.pal(7,"Greens")
ggplot(education_and_poverty_map, aes(x=long, y = lat, group = group, 
                                      map_id = region, fill = PCTPOVALL_2013)) +
  geom_map(map = education_and_poverty_map, color = "black" ) +
  #scale_fill_gradientn(colors=colors) +
  coord_map("polyconic") +
  ggtitle("Fig #: ")

```


