---
title: "Using crosstab"
author: "Elin Waring"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using crosstab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, message=FALSE}
library(lehmansociology)
addhealth<-droplevels.data.frame(addhealth)
```

This document describes how to use the ` crosstab() ` function from lehmansociology.  
This function is designed to produce useful crosstabs that can have 
raw frequencies, row, colum and total percents and either numeric or 
percent marginals

## Basic use

By default the function produces a table of raw frequencies with
a bottom row containing the column N values.  Note that the 
data= can be dropped if the data set name is in the first position.
The row.vars and col.vars list variables from the data set by name, and
the name must be in quotation marks.  

You must have both row and column variables. For a single variable, 
use frequency(). 

```{r}

crosstab(addhealth, row.vars = "happywschool", col.vars="sex")

```

You can use additional parameters to modify the table.

For example, let's make a table with column percents and a marginal
column with the percents for the whole sample.


```{r}

crosstab(data = addhealth, row.vars = "happywschool", col.vars="sex",
         format = "column_percent", row.margin.format = "percent")

```

If you would like to have row percents (which is not normal practice
in sociology but is in other disciplines) you can change the format to 
` row_percents`.  If you would like total percents (percent of the whole
sample that is in a given cell) use total_percent. If you would like
frequencies instead of percents in the total column change the
row.margin.format to "freq".


```{r}

crosstab(data = addhealth, row.vars = "happywschool", col.vars="sex",
         format = "total_percent", row.margin.format = "freq")

```

You can also add a title using the title="" parameter. You can change
the Total N row to show the percent in each column by adding 
` col.margin.format="percent" `  or eliminate that row by using ` "none" `.


```{r}

 crosstab(data = addhealth, row.vars = "happywschool", col.vars="sex",
          format = "column_percent", row.margin.format = "percent",
          title = "Feeling Happy with School by Gender", 
          col.margin.format = "percent")

```

You can get a summary including Chi Square by using the summary function.

```{r}

result<- crosstab(data = addhealth, row.vars = "happywschool", col.vars="sex",
         format = "column_percent", row.margin.format = "percent",
         title = "Feeling Happy with School by Gender",
         col.margin.format = "percent")

summary(result)

```

By default the parameter pretty.print = FALSE, but when you are ready to knit you may want to change it to
TRUE and use pander to print. If you run this without knitting it will look a bit strange if 
you have multiple lines, but
once you knit it will look nicer. Of course you can change many options on this, such as justification
and whether to use linebreaks. Overall, it is a more professional approach to displaying your results.
Notice that you need to add the $tab when using `pander()` to print.
```{r}

library(pander)
t<-crosstab(addhealth, row.vars = "happywschool", col.vars = c("sex", "raceeth"), pretty.print = TRUE)
t
pander(t$tab, keep.line.breaks = TRUE, justify = "right")

```
You can learn about many options for pander at the pander website. 

## Advanced

The function returns an object of class crosstab.

The object is a list containing:  

tabf   

A table object with the raw frequencies.

*tab*  

The created table that is printed when you run crosstab().

*n*    

The number of observations in the table.
 
*margins*   

A list of freqency and proportion margins for the table.

*title*  

This is the user supplied title or an empty string (default).

Once you have the object you can use these separately or modify them 
using normal R procedures. 
