---
title: "Using Frequency()"
author: "Elin Waring"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, message=FALSE}
library(lehmansociology)
```

This document describes how to use the freqency() function from lehmansociology.  
This function is designed to produce useful frequency, percent and
cumulative data distributions in tabular form for variables treated as
factors or ordinal factors. 

## Basic use

By default the function produces a table containing the frequency and
percent. 

```{r}
frequency(chickwts$feed)

```

You can set freq = FALSE or percent = FALSE to eliminate the respective 
columns from the results

```{r}

frequency(chickwts$feed, freq = FALSE)

```

If apporopriate, you can also request the cumulative frequencies and percents
using cumulative.freq = TRUE and cumulative.percent = TRUE. Note that 
cumulative columns will have a blank Total row.  

```{r}

frequency(chickwts$weight, cumulative.percent = TRUE)

```

Add a title to the table using the title parameter.

```{r}

frequency(chickwts$feed, title = "Feed types")

```

Handling of missing data is done using the useNA parameter. By default 
useNA = "no" which means that missing data are excluded from the results.
useNA = "ifany" will include missing values if they are present. useNA =
"always" will add an NA row to the table whether or not there is any
missing data.

## Printing

Printing is what controls what data R displays either when you type `x` 
or `print(x)` and how it looks.   

The function `print.frequencytable` lets you print the table with different options.  
You only need to use this if you want to do something other than the defaults.  

It includes the same parameters as `print.table()` from base r but by
default row names are not printed and the values of the variable are left
justified.  Use `row.names = TRUE` to print
the names or numbers. Use  `right = TRUE` to right justify the values.   

## Advanced

The function returns an object of class frequencytable.

The object is a list containing:  

table  
This is the whole table as a dataframe. 

frequencies 
This is the table class for the variable. That is, it is the result from
using `table(x)` with the value of useNA from the `frequency()` call.  

proportions  
This is the table class result for `prop.table(table(x))`.  

cum_prop  
This is the cumulative proportions as integers. Effectively this is
`cumsum(prop.table(table(x)))`.  

cum_freq  
This is the cumulative frequencies as integers. Effectively `cumsum(table(x))`.

totrow  

This is the row displayed at the bottom. The first element contains the
string "Total".  Cumulative totals (totrow[3] and totrow[4]) are empty
strings.

title

This is the supplied title or an empty string (default).

Once you have the object you can use these separately or modify them 
using normal R procedures. 
